---
title: "Developing R packages"
subtitle: "OECD Stats Day 2023"
author: 
  - name: "María Paula Caldas"
    affiliation: 
      - "Economics Department"
      - "R/Python/Algobank CoP"
# code-line-numbers: false
format: 
  revealjs:
    incremental: false
    theme: [simple, custom.scss]
    css: custom.css
    slide-number: true
    # footer: "<https://mpaulacaldas.github.io/oecd-r-pkg-dev>"
    # logo: www/SYMBOL_20cm.png
workshop:
  slides: https://mpaulacaldas.github.io/oecd-r-pkg-dev
  project: https://posit.cloud/content/7235458
execute: 
  eval: false
  echo: true
knitr: 
  opts_chunk: 
    collapse: true
    comment: "#>" 
---

## Useful links

\

{{< fa link >}} Slides

<https://mpaulacaldas.github.io/oecd-r-pkg-dev>

{{< fa folder-open >}} Posit Cloud Project

<https://posit.cloud/content/7235458>


# Introduction {background-color="#2a2e45ff"}

Motivation and prerequisites

## What is an R package? 

A package bundles together **code**, **data**, and **documentation** in a 
format that is **easy to share with others**

```{r}
#| code-line-numbers: "|1-2|6-8|10-12|14-16"
#| label: ggplot2-example
#| eval: false
# R makes it easy to install and use packages from CRAN or other repositories
install.packages("ggplot2")

library(ggplot2)

# Some packages contain data, which can be documented
data(package = "ggplot2")
?ggplot2::diamonds

# Beyond object documentation, packages can include short articles to 
# describe broader functionality
vignette("ggplot2-specs")

ggplot(diamonds, aes(x, y)) + 
  geom_bin_2d(show.legend = FALSE) +
  scale_y_log10()
```

## Why package your R code?

- Makes it easy to share your code with others on your team or across the 
  Organisation (add examples from positconf, internal from the cop)
- Adopting a package infrastructure makes it easy for you as a developer to 
  create more robust code (e.g. versioning, unit tests, declaring dependencies)
- Packages also make it easier for others to use your code (e.g. function 
  documentation, vignettes, limit namespace conflicts)

## Pre-requisites

- be willing!
- know R, enough to know how to create a function (must-have)
- know how to use Git (nice-to-have)
- know markdown (nice-to-have)


# Let's create a package {background-color="#2a2e45ff"  background-image="https://usethis.r-lib.org/logo.png" background-size="150px" background-position="1050px 50px"}

\ 

{{< fa folder-open >}} Posit Cloud Project

<https://posit.cloud/content/7235458>

## [1]{.circle} Set-up the basic infrastructure

Let's load `{usethis}` and `{devtools}` into your global environment. These are 
**workflow** packages. They are used interactively during package development.

```{r}
library(usethis)
library(devtools)

create_package("location-i-want/mypackage")
```

::: {.incremental}

- What happens when you run `create_package()`?
- What files do you see?
- Open the `DESCRIPTION` file and edit some fields.
- Run `devtools::check()`. What do you see?

:::

## [2]{.circle} Create a function, and use it!

\

::: {.incremental}

- Open a file to write your function.
  ```{r}
  use_r("name-of-your-file")
  ```

- Write a small, simple function.
  ```{r}
  #| code-fold: true
  #| code-summary: "If you are low on inspiration..."
  year_progress <- function(date, is_leap_year = FALSE) {
    nominator   <- as.numeric(format(date, "%j"))
    denominator <- if (is_leap_year) 366 else 365
    share <- round(nominator * 100 / denominator)
    message(share, "% of the year is done!")
  }
  ```
- When you are done, write `devtools::load_all()` in the console and try 
  to use your function. What happens?

:::

## [3]{.circle} Document
- Insert roxygen
- document()

## [📝 NEW CONCEPTS]{.tag}{background-color="#a499beff"}
### Internal and exported functions

- Optional: add a dependency

## [4]{.circle} Test
- use_test()
- check()

## [5]{.circle} Share
- use_git() + use_algobank()
- install()

## [📝 NEW CONCEPTS]{.tag}{background-color="#a499beff"}
### Source vs. binary

## Recap

# Lessons {background-color="#2a2e45ff"}

## For developers

- Read the book
- Use a good style
- UTF-8!
- use_clean_slate()
- dev-script
- OECD: local version of R, local folder

## For managers

- Don't compound training curves
- Pay for training, encourage training or give time for training
- Align incentives
- The larger the number of users, the higher the engineering bar
- Think of governance and maintenance

## Lessons from the OECD 

- Eco Data Platform
- Community of Practice, oecdverse
